### 2014-03-28 patchy-selection-paper.tex
.

Migration--selection balance ensures that 
there will be some number of $B$ alleles present in the regions where they are disadvantageous,
but only very few, and rarely, far away from the patch where $B$ is established.
Denote the expected frequency of allele $B$ at displacement $x$ relative to the patch by $q(x)$.
Following \citet{slatkin1973geneflow}, one can write down a differential equation to which $q(x)$ is the solution,
which in one dimension can be solved to give $q(x)$ in terms of Jacobi elliptic functions (see Appendix \ref{apx:elliptic_integrals}).
More general, asymptotics of the solution implies that for large $r$ the mean occupation frequency decays exponentially, with a square-root correction in two dimensions (see Appendix \ref{apx:asymptotics}): 
\begin{align} \label{eqn:eqfreq}
    q(x) \approx C |x|^{-(d-1)/2} \exp( - |x| \sqrt{2 s_m} / \sigma) \qquad \text{for large $|x|$},
\end{align}
where $d$ is the dimension ($d=1$ or $2$), and $C$ is a constant depending on the geographic shape of the populations and the selection coefficients. 
In practice we take $C=1/2$,
since from simulations, $C$ does not seem to vary from this by more than a factor of 2 or so (see Appendix \ref{apx:patchy_sims}).

@@

To make the argument, we decompose the genealogy of $B$ alleles into families in the following way.
First, consider the genealogy of all $B$ alleles that were alive at any time outside of the original patch.  
This is a collection of trees, each rooted at an allele living outside the patch whose parent was born inside the patch.
Next, fix some intermediate distance $r_0$ from the established patch,
and erase from the genealogy every allele that has no ancestors living further away than $r_0$ to the patch.
This fragments the original set of trees into a collection of smaller trees that relate to each other all $B$ alleles living outside of $r_0$ at any time,
and some living inside of $r_0$;
we refer to these trees as ``families''.
If $r_0$ is large enough that there are not too many families
and interactions between family members are weak,
then these families can be approximated by a collection of independent spatial branching processes
whose members are killed (actually, ignored) if they enter the original patch.
(This can be made formal in the limit of large population density, also taking $r_0$ large enough that the chance of reaching the original patch is small.)
This decomposition is illustrated in~\ref{fig:sim_snapshots}, and later in figure~\ref{fig:branching_decomp}.
In terms of these families, the equilibrium frequency above is then
\begin{align}
    \label{eqn:gestalt_q}
    \begin{split}
        q(x) &= \text{ ( outflux of families ) } \times \text{ ( mean occupation density of a family at $x$ ) } ,
\end{split}
\end{align}
and the quantity we wish to compute,
the effective rate of at which \emph{families of} migrant $B$ alleles establish in a new patch at distance $r$, 
is
\begin{align}
    \label{eqn:gestalt_migrate}
    \begin{split}
        \migrate(x) &= \text{ ( outflux of families ) } \times \text{ ( probability a family establishes in patch at $x$ ) }
    \end{split}
\end{align}
These quantities should not depend on $r_0$, of course.
All terms except the ``outflux of families'' are calculations with spatial branching processes;
and we can solve for this outflux using the established form for $q(r)$ in \eqref{eqn:gestalt_q}.
Note that $\migrate$ reflects the mean number of establishing migrant families per unit time,  rather than individuals.

@@

First, we compute the second term in equation \eqref{eqn:gestalt_q}, (mean occupation density of a family at $x$),
by which we mean the expected total number of offspring of a single family that ever live at distance $r$ from the patch.
In two dimensions, this depends on where the family began, 
and right-hand side of equation \eqref{eqn:gestalt_q} should be integrated over this starting point.
Here we'll write $x_0$ for the starting point, and $u(x_0,x)$ for the mean occupation density at $x$.
Concretely, suppose that $Z$ is a spatial branching process as above, started with a single individual at $x_0$,
defined now so that $Z_t(A)$ is the number of descendant individuals alive at time $t$ living in the set $A$.
The mean occupation density of $Z$ is then defined to be the density of $\E[\int_0^t Z_t dt]$,
i.e.\ the smooth function $u(x_0,x)$ that, integrated over a set $A$, 
gives the expected total number of individuals ever living in $A$:
\begin{align} \label{eqn:occupation_density_defn}
    \E\left[ \int_0^\infty Z_t(A) dt \right] = \int_A u(x_0,x) dx .
\end{align}
Since a family has on average $(1-s_m)^t$ members in generation $t$,
and the marginal distribution of any one surviving individual is approximately Gaussian with variance $t \sigma^2$,
exchanging the expectation and the integral shows that
the density $u(x_0,x)$ is equal to the Green's function for Brownian motion killed at rate $1-s_m$,
\begin{align} \label{eqn:occupation_density_solution}
    u(x_0,x) = \int_0^\infty (1-s_m)^t p_{\sigma^2 t}(x_0,x) dt
\end{align}
where $p_t(x,y) = (2 \pi t)^{-d/2} \exp(-\|x-y\|^2/2t)$ is the transition density for standard Brownian motion.
This is known exactly; the solution is

@@

Now, we can obtain the first term in equation \eqref{eqn:gestalt_q}, the (outflux of families).
Define $\gamma$ to be this outflux; 
i.e.\ the mean number of individuals per unit time that are born carrying the $B$ allele closer than $r_0$ to the original patch
but migrate to a larger distance.
Then for $r>r_0$, the mean occupation density $q(r)$ is equal to $\gamma$ 
multiplied by the mean occupation density of the family descending from such a migrant,
which we denote $u(r)$.
Denote by $Z_t$ the number of individuals such a family has after time $t$,
and the locations of these by $X_1(t), \ldots, X_{Z_t}(t)$.
Then this mean occupation density, at $x$, is the expected number of individuals ever present at $x$. 
Since the expected number of individuals present at time $t$ is $e^{\alpha t}$
and the location of each of these individuals is marginally distributed as a  Gaussian with variance $\sigma^2 t$,
we can write this expectation as
\begin{align}
    u(x) &= \int_0^t \E[ \#\{ k : \; X_k(t) \in dx \} ] dt \\
    &= \int_0^t e^{\alpha t} \frac{1}{\sqrt{2 \pi t \sigma^2}} e^{-\frac{x^2}{2t\sigma^2}} dt \\
    &= \frac{ 1 }{ \sigma \sqrt{2 \alpha} } \exp\left( - \frac{ x \sqrt{2\alpha} }{ \sigma } \right) 
\end{align}
\plr{check that}
The integral is evaluated in Appendix~\ref{apx:some_integrals}.
\plr{Also examine non-dependence on $r_0$.}

@@

In one dimension, we are done (see section~\ref{apx:elliptic_integrals});
in higher dimensions, we are more interested in the mean frequency at a given distance $r$ from a patch.
If we take a radially symmetric patch centered at the origin (so $\gamma$ only depends on $r$), 
and let $\xi(t,r)$ denote the mean occupation frequency at distance $r$,
then the polar form of the Laplacian in $d$ dimensions gives us that \eqref{eqn:derived_fkpp} is
\begin{align}
  \label{eqn:radial_fkpp}
  \partial_t \xi(t,r) = \mu \left( \frac{\sigma^2}{2} \partial_{r}^2 \xi(t,r) + \sigma^2\frac{d-1}{2r} \partial_r \xi(t,r) + \gamma(r) \xi(t,r) (1-\xi(t,r)) \right) .
\end{align}
However, the second term goes to zero as $r \to \infty$,
leading to the same asymptotic behavior for the tail \plr{CITE SOMETHING}.

%%%%%%%
\section{The exact solution to the equilibrium frequency}
\label{apx:elliptic_integrals}

@@

%%%%%%%%%%%
\section{Patchy selection} 
\label{apx:asymptotics}

The general equation XXX is not analytically solvable,
but the theory in \citet{wasow1965asymptotic}
lets us conclude that XXX

>>>>

Migration--selection balance ensures that 
there will be some number of $B$ alleles present in the regions where they are disadvantageous,
but only very few, and rarely, far away from the patch where $B$ is established.
Denote the expected frequency of allele $B$ at displacement $x$ relative to the patch by $q(x)$.
Following \citet{slatkin1973geneflow}, one can write down a differential equation to which $q(x)$ is the solution,
which in one dimension can be solved to give $q(x)$ in terms of Jacobi elliptic functions (see Appendix \ref{apx:elliptic_integrals}).
More general, asymptotics of the solution implies that for large $r$ the mean occupation frequency decays exponentially, with a square-root correction in two dimensions (see Appendix \ref{apx:asymptotics}): 
\begin{align} \label{eqn:eqfreq}
    q(x) \approx C (|x|-r_0)^{-(d-1)/2} \exp( - |x| \sqrt{2 s_m} / \sigma) \qquad \text{for large $|x|$},
\end{align}
where $d$ is the dimension ($d=1$ or $2$), and $C$ and $r_0$ are constants depending on the geographic shape of the populations and the selection coefficients
(but note that $r_0$ is only necessary in $d=2$).
In applications we fit $C$ and $r_0$ to the data;
to get concrete numbers from theory we take $C=1$ and $r_0=0$ if necessary.
From simulations, $C$ seems to be close to 1 in practice (see Appendix \ref{apx:patchy_sims})
and XXX $r_0$.

@@

To make the argument, we decompose the genealogy of $B$ alleles into families in the following way.
First, consider the genealogy of all $B$ alleles that were alive at any time outside of the original patch.  
This is a collection of trees, each rooted at an allele living outside the patch whose parent was born inside the patch.
Next, fix some intermediate distance $r_0$ from the established patch,
and erase from the genealogy every allele that has no ancestors living further away than $r_0$ to the patch.
This fragments the original set of trees into a collection of smaller trees that relate to each other all $B$ alleles living outside of $r_0$ at any time,
and some living inside of $r_0$;
we refer to these trees as ``families''.
If $r_0$ is large enough that there are not too many families
and interactions between family members are weak,
then these families can be approximated by a collection of independent spatial branching processes
whose members are killed (actually, ignored) if they enter the original patch.
(This can be made formal in the limit of large population density, also taking $r_0$ large enough that the chance of reaching the original patch is small.)
This decomposition is illustrated in~\ref{fig:sim_snapshots}, and later in figure~\ref{fig:branching_decomp}.
In terms of these families, the equilibrium frequency above is then
\begin{align}
    \label{eqn:gestalt_q}
    \begin{split}
        q(x) &= \text{ ( outflux of families ) } \times \text{ ( mean occupation density of a family at $x$ ) } ,
\end{split}
\end{align}
and the quantity we wish to compute,
the effective rate of at which \emph{families of} migrant $B$ alleles establish in a new patch at distance $r$, 
is
\begin{align}
    \label{eqn:gestalt_migrate}
    \begin{split}
        \migrate(x) &= \text{ ( outflux of families ) } \times \text{ ( probability a family establishes in patch at $x$ ) }
    \end{split}
\end{align}
The quantities on the right-hand side depend implicitly on $r_0$.
All terms except the ``outflux of families'' are calculations with spatial branching processes;
and we can solve for this outflux using the established form for $q(r)$ in \eqref{eqn:gestalt_q}.
Note that $\migrate$ reflects the mean number of establishing migrant families per unit time,  rather than individuals.

@@

First, we compute the second term in equation \eqref{eqn:gestalt_q}, (mean occupation density of a family at $x$),
by which we mean the expected total number of offspring of a single family that ever live at distance $r$ from the patch.
In two dimensions, this depends on where the family began, 
and right-hand side of equation \eqref{eqn:gestalt_q} should be integrated over this starting point.
Here we'll write $x_0$ for the starting point, and $u(x_0,x)$ for the mean occupation density at $x$.
Concretely, suppose that $Z$ is a spatial branching process as above, started with a single individual at $x_0$,
defined now so that $Z_t(A)$ is the number of descendant individuals alive at time $t$ living in the set $A$.
The mean occupation density of $Z$ is then defined to be the density of $\E[\int_0^t Z_t dt]$,
i.e.\ the smooth function $u(x_0,x)$ that, integrated over a set $A$, 
gives the expected total number of individuals ever living in $A$:
\begin{align} \label{eqn:occupation_density_defn}
    \E\left[ \int_0^\infty Z_t(A) dt \right] = \int_A u(x_0,x) dx .
\end{align}
Since a family has on average $(1-s_m)^t$ members in generation $t$,
and the marginal distribution of any one surviving individual is approximately Gaussian with variance $t \sigma^2$,
exchanging the expectation and the integral shows that
the density $u(x_0,x)$ is equal to the Green's function for Brownian motion killed at rate $1-s_m$,
\begin{align} \label{eqn:occupation_density_solution}
    u(x_0,x) = \int_0^\infty (1-s_m)^t p_{\sigma^2 t}(x_0,x) dt
\end{align}
where $p_t(x,y) = (2 \pi t)^{-d/2} \exp(-\|x-y\|^2/2t)$ is the transition density for standard Brownian motion.
The integral can be expressed in terms of modified Bessel functions:
using \citet{gradshteyn2007table} (3.471.9 and 8.451.6), the mean occupation density is,
with $\alpha = -\log(1-s_m)$ and $d$ the dimension,
\begin{align}
    u(x_0,x) &= \frac{1}{\sigma^d \pi}\left(\sqrt{ \frac{\|x-x_0\|^2}{2\alpha\sigma^2} }\right)^{(d+2)/2} K_{(d+2)/2}\left( \sqrt{2\alpha} \frac{\|x-x_0\|}{\sigma} \right) \\
             &= \frac{1}{\sigma^d \sqrt{2\pi}} \left(\sqrt{ \frac{\|x-x_0\|^2}{2\alpha\sigma^2} }\right)^{d/2} \exp\left(- \sqrt{2\alpha} \frac{\|x-x_0\|}{\sigma} \right)+ O(\sigma/\|x-x_0\|) . \label{eqn:occupation_asymptotics}
\end{align}
Since we only have asymptotics for the equilibrium frequency, 
we will only use the latter (asymptotic) form.

@@

Now, we can obtain the first term in equation \eqref{eqn:gestalt_q}, the (outflux of families).
Define $\gamma(r_0)$ to be this outflux; 
i.e.\ the mean number of individuals per unit time that are born carrying the $B$ allele closer than $r_0$ to the original patch
but migrate to a larger distance.
Then for any point $x$ further than $r_0$ from the patch, 
the mean occupation density $q(x)$ is equal to the integral of $\gamma$ over all locations $x_0$ at distance $r_0$ from the patch,
multiplied by the mean occupation density at $x$ for migrant families beginning at at location.
This is a more precise statement of the decomposition of the equilibrium frequency from equation \eqref{eqn:gestalt_q}.
However, this integral is not tractable; if we assume that $r_0$ is small relative to $\|x\|$, then we get 
\begin{align}
    q(x) &= \int_{B(r_0)} u(y,x) \gamma(|y|) dy \\
         &\simeq \gamma(r_0) 2 (\pi r_0)^{d-1} u(0,x) .
\end{align}
Now, using equations \eqref{eqn:eqfreq} and \eqref{eqn:occupation_asymptotics}, we get that
\begin{align}
    \gamma(r_0) \simeq 
\end{align}

Denote by $Z_t$ the number of individuals such a family has after time $t$,
and the locations of these by $X_1(t), \ldots, X_{Z_t}(t)$.
Then this mean occupation density, at $x$, is the expected number of individuals ever present at $x$. 
Since the expected number of individuals present at time $t$ is $e^{\alpha t}$
and the location of each of these individuals is marginally distributed as a  Gaussian with variance $\sigma^2 t$,
we can write this expectation as
\begin{align}
    u(x) &= \int_0^t \E[ \#\{ k : \; X_k(t) \in dx \} ] dt \\
    &= \int_0^t e^{\alpha t} \frac{1}{\sqrt{2 \pi t \sigma^2}} e^{-\frac{x^2}{2t\sigma^2}} dt \\
    &= \frac{ 1 }{ \sigma \sqrt{2 \alpha} } \exp\left( - \frac{ x \sqrt{2\alpha} }{ \sigma } \right) 
\end{align}
\plr{check that}
The integral is evaluated in Appendix~\ref{apx:some_integrals}.
\plr{Also examine non-dependence on $r_0$.}

@@

In one dimension, we are done (see section~\ref{apx:elliptic_integrals});
in higher dimensions, we are more interested in the mean frequency at a given distance $r$ from a patch.
If we take a radially symmetric patch centered at the origin (so $\gamma$ only depends on $r$), 
and let $\xi(t,r)$ denote the mean occupation frequency at distance $r$,
then the polar form of the Laplacian in $d$ dimensions gives us that \eqref{eqn:derived_fkpp} is
\begin{align}
  \label{eqn:radial_fkpp}
  \partial_t \xi(t,r) = \mu \left( \frac{\sigma^2}{2} \partial_{r}^2 \xi(t,r) + \sigma^2\frac{d-1}{2r} \partial_r \xi(t,r) + \gamma(r) \xi(t,r) (1-\xi(t,r)) \right) .
\end{align}

%%%%%%%%%%%
\subsection{Asymptotic solution for the equilibrium frequency} 
\label{apx:asymptotics}

If $q(r)$ is a radially symmetric equilibrium frequency and $\xi(t,x)=q(|x|)$,
then $\partial_t \xi(t,x) = 0$.
So, using \eqref{eqn:radial_fkpp}, a radially symmetric solution 
with $\gamma(r) = -s < 0$ for all $r>r_0$ 
solves, for $r_0 < r < \infty$,
\begin{align}
    \partial_r^2 q(r) + \frac{d-1}{r} \partial_r q(r) - \frac{2s}{\sigma^2} q(r) (1-q(r)) = 0 \\
    \lim_{r \to \infty} q(r) = \lim_{r \to \infty} \partial_r q(r) = 0 \qquad 
    0 < q(r) < 1 
\end{align}
Since $q(r) \to 0$ as $r \to \infty$, so $q(r) (1-q(r)) \approx q(r)$,
XXX theory in \citet{wasow1965asymptotic} XXX
implies that the true equilibrium frequency $q$ is close, for large $r$, to the solution to
\begin{align}
    \partial_r^2 u(r) + \frac{d-1}{r} \partial_r u(r) - \frac{2s}{\sigma^2} u(r) = 0  \label{eqn:bessel} \\
    \lim_{r \to \infty} u(r) = \lim_{r \to \infty} \partial_r u(r) = 0 \qquad  .
\end{align}
This has general solution given by a modified Bessel function:
using \citet{gradshteyn2007table} 8.494.8,
the general solution is
\begin{align}
    u(r) = C' (r-r_0)^{d/2} K_{d/2} \left( (r-r_0) \sqrt{2s}/\sigma \right) ,
\end{align}
where $C'$ and $r_0$ are chosen to match boundary conditions.
Asymptotics of Bessel functions \citep[][8.451.6]{gradshteyn2007table} then imply that
\begin{align}
    q(r) \simeq C (r-r_0)^{(d-1)/2} \exp \left( -r \sqrt{2s}/\sigma \right) ,
\end{align}
where $C$ is a different constant.

%%%%%%%
\subsection{The exact solution to the equilibrium frequency}
\label{apx:elliptic_integrals}

@@
